---
title: "MOFA Factor-Feature correlations"
subtitle: "MOFA model which includes metabolites"
author: "Paul Heinrich"
output: 
  html_document:
    toc: TRUE
---

# Notes

Metabolite data was included in the model. No scaling of omics views to unit variance. 
Fitting was performed with medium convergence and 10 factors. No subsets of view features were
selected, so views have a substantial inequality in the number of features (especially
metabolites, with only around 30 features).

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}

library("here")
library("ggplot2")
library("ggpubr")
library("rstatix")
library("stringr")
library("patchwork")
library("MOFA2")
library("knitr")
library("dplyr")
library("pheatmap")
library("RColorBrewer")
library("grid")
library("gridExtra")
library("WriteXLS")

here::i_am("analysis/hep62703_2022-10-19_meta_F-10_C-medium_no_scaling/reports/mofa_plots_factor_feature_corr.Rmd")

source(here("functions/load_mofa_tmp_copy.R"))

save_data = FALSE

today <- Sys.Date()

```

```{r load_data, echo=FALSE, message = FALSE}

analysis_path <- here("analysis/hep62703_2022-10-19_meta_F-10_C-medium_no_scaling")

dir_model <- file.path(analysis_path, "data", "models")
name_model <- "model_metabolites_medium_F10.hdf5"

plot_path <- file.path(analysis_path, "img")
table_path <- file.path(analysis_path, "tables")

model <- load_mofa_tmp_copy(
    file_directory = dir_model,
    file_name = name_model,
    name_tmp_folder = "p107_hp_mofa"
)

taxonomic_mapping <- list()
mapping_views <- c("16S", "ITS")

for (view in mapping_views) {
    filename <- paste0("taxonomic_mapping_", view, ".csv")
    taxonomic_mapping[[view]] <- read.csv(
        here("metadata/taxonomic_classification", filename),
        header=TRUE,
        row.names = 1
    )
    mapping_vec <- taxonomic_mapping[[view]][["Genus"]]
    names(mapping_vec) <- rownames(taxonomic_mapping[[view]])
    taxonomic_mapping[[view]] <- mapping_vec
}

```

```{r get_model_data, echo=FALSE}

views <- views_names(model)
metadata <- samples_metadata(model)
factor_values <- get_factors(model, factors = "all")$group1
factor_weights <- get_weights(model)
model_input <- get_data(model)

metadata[["pat_GvHD"]] <- as.factor(metadata[["pat_GvHD"]])
metadata[["GvHD_patient"]] <- as.factor(metadata[["GvHD_patient"]])
metadata[["pat_TRM_1yr"]] <- as.factor(metadata[["pat_TRM_1yr"]])

metadata_factors <- merge(
    x = metadata,
    y = factor_values,
    by.x = "sample",
    by.y = "row.names"
)

timepoints_to_exclude <- c(
    "allo",
    "idx"
)

metadata_factors_subset <- metadata_factors[
    !(metadata_factors[["pre_post_allo_SCT"]] %in% timepoints_to_exclude),
]

metadata_factors_subset[["pre_post_allo_SCT"]] <- as.factor(
    metadata_factors_subset[["pre_post_allo_SCT"]]
)

factor_names <- colnames(metadata_factors_subset)[
    grepl("Factor", colnames(metadata_factors_subset))
]

factors <- as.numeric(str_extract(factor_names, "[0-9]+$"))
names(factors) <- factor_names

clinical_factors <- c("GvHD_patient", "pat_TRM_1yr")
clinical_factors_mapping <- c("GvHD", "1-year-TRM")
names(clinical_factors_mapping) <- clinical_factors
pre_post_values <- unique(metadata_factors_subset[["pre_post_allo_SCT"]])

relevantFactors <- factors[c(1, 3, 4)]
relevantViews <- views[c(1, 3, 4)]

```

```{r factor_feature_correlations, echo=FALSE}

factor_feature_corr <- list()
factor_feature_corr_p <- list()
p_vals <- c()
corr_coefs <- c()

for (factor in names(relevantFactors)) {
    factor_feature_corr[[factor]] <- list()
    factor_feature_corr_p[[factor]] <- list()
    curr_factor <- factor_values[, factor]
    for (view in views) {
        curr_weights <- factor_weights[[view]][, factor]
        curr_weights <- curr_weights[order(abs(curr_weights), decreasing=TRUE)]

        corr <- lapply(
            names(curr_weights),
            function(x) cor.test(
                x = curr_factor,
                y = model_input[[view]]$group1[x, ]
            )
        )
        names(corr) <- names(curr_weights)
    
        factor_feature_corr[[factor]][[view]] <- sapply(
            names(corr),
            function(x) unname(corr[[x]]$estimate)
        )
        factor_feature_corr_p[[factor]][[view]] <- sapply(
            names(corr),
            function(x) unname(corr[[x]]$p.value)
        )

        curr_p_vals <- factor_feature_corr_p[[factor]][[view]]
        names(curr_p_vals) <- paste0(factor, "_", view, "_", names(curr_p_vals))

        p_vals <- c(p_vals, curr_p_vals)

        curr_corr_coefs <- factor_feature_corr[[factor]][[view]]
        names(curr_corr_coefs) <- paste0(factor, "_", view, "_", names(curr_corr_coefs))

        corr_coefs <- c(corr_coefs, curr_corr_coefs)
    }
}

p_vals_adjusted <- p.adjust(p_vals, method = "BH")

```

# Factor-feature correlation heatmaps

Correlation of factor values with top weight features.

```{r factor_feature_heatmaps, echo=FALSE}

n_weights <- 15
factor_feature_heatmaps <- list()

for (factor in names(relevantFactors)) {
    factor_feature_heatmaps[[factor]] <- list()
    for (view in views) {
        feature_ids <- as.character(
            names(factor_feature_corr[[factor]][[view]])[1:n_weights]
        )
        feature_ids <- str_remove(feature_ids, paste0("_", view, "$"))

        if(view %in% names(taxonomic_mapping)) {
            feature_names <- taxonomic_mapping[[view]][feature_ids]
        } else {
            feature_names <- feature_ids
        }

        mat <- matrix(factor_feature_corr[[factor]][[view]][1:n_weights], nrow = 1)
        colnames(mat) <- feature_names

        hm <- pheatmap(
            t(mat),
            cluster_rows = FALSE,
            cluster_cols = FALSE,
            breaks=seq(-1, 1, length.out=101),
            cellwidth = 40,
            legend = FALSE,
            silent = TRUE
        )
        factor_feature_heatmaps[[factor]][[view]] <- hm$gtable
    }
}

print("Factor 1, pheatmap")

grid.arrange(arrangeGrob(
    grobs=factor_feature_heatmaps$Factor1[c("16S", "virome", "metabolites")], ncol=3
))

print("Factor 3, pheatmap")

grid.arrange(arrangeGrob(
    grobs=factor_feature_heatmaps$Factor3[c("16S", "virome", "metabolites")], ncol=3
))

```

```{r factor_feature_heatmaps_gg, echo=FALSE}

factor_feature_heatmaps_gg <- list()

for (factor in names(relevantFactors)) {
    factor_feature_heatmaps_gg[[factor]] <- list()
    for (view in views) {
        feature_ids <- as.character(
            names(factor_feature_corr[[factor]][[view]])[1:n_weights]
        )
        feature_ids <- str_remove(feature_ids, paste0("_", view, "$"))

        if(view %in% names(taxonomic_mapping)) {
            feature_names <- taxonomic_mapping[[view]][feature_ids]
        } else {
            feature_names <- feature_ids
        }

        curr_data <- data.frame(
            "view" = view,
            "feature_id" = feature_ids,
            "correlation" = factor_feature_corr[[factor]][[view]][1:n_weights],
            "feature" = feature_names
        )

        hm <- ggplot(curr_data, aes(view, feature_id, fill = correlation)) + 
            geom_tile() +
            scale_fill_gradient2(
                low = "blue", mid = "white", high = "red",
                limits = c(-1, 1)
            ) + labs(
                x = NULL,
                y = NULL
            )
        
        hm <- hm + scale_y_discrete(
                breaks=hm$data[["feature_id"]],
                labels=hm$data[["feature"]]
        )

        factor_feature_heatmaps_gg[[factor]][[view]] <- hm
    }
}

print("Factor 1, ggplot")

factor_feature_heatmaps_gg[["Factor1"]][["16S"]] +
factor_feature_heatmaps_gg[["Factor1"]][["virome"]] +
factor_feature_heatmaps_gg[["Factor1"]][["metabolites"]] +
plot_layout(guides = 'collect')

print("Factor 3, ggplot")

factor_feature_heatmaps_gg[["Factor3"]][["16S"]] +
factor_feature_heatmaps_gg[["Factor3"]][["virome"]] +
factor_feature_heatmaps_gg[["Factor3"]][["metabolites"]] +
plot_layout(guides = 'collect')

```

# Fraction of significant factor-feature associations

```{r fraction_sig_associations, echo=FALSE}

sig_threshold <- 0.05

total_num_fractions <- length(relevantFactors)*length(views)
sig_assoc <- c()
factor_id <- c()
view_id <- c()

for (factor in names(relevantFactors)) {
    for (view in views) {
        factor_id <- c(factor_id, factor)
        view_id <- c(view_id, view)

        curr_subset <- p_vals_adjusted[
            grep(paste0("^", factor, "_", view), names(p_vals_adjusted))
        ]
        curr_frac <- sum(curr_subset <= sig_threshold)/length(curr_subset)

        sig_assoc <- c(sig_assoc, curr_frac)
    }
}

frac_sig_associations <- data.frame(
    "factor" = factor_id,
    "view" = factor(view_id, levels = rev(views)),
    "fraction_sig" = sig_assoc
)

```

```{r fraction_sig_associations_plots, echo=FALSE}

fraction_sig_assoc_plot <- ggbarplot(
    data = frac_sig_associations,
    x = "factor",
    y = "fraction_sig",
    fill = "view",
    xlab = FALSE,
    ylab = "Fraction of significant associations"
) + geom_text(
    aes(label = round(fraction_sig, 2)),
    position = position_stack(vjust = 0.5),
    size = 4
) + scale_fill_manual(
    values = rev(c("#F9F2ED", "#3AB0FF", "#FFB562", "#F87474")),
    name = NULL,
    breaks = c("16S", "ITS", "virome", "metabolites"),
    labels = c("Bacteriome", "Fungome", "Virome", "Metabolites"),
) + scale_x_discrete(
    breaks = c("Factor1", "Factor3", "Factor4"),
    labels = c("Factor 1", "Factor 3", "Factor 4")
) + theme(
    legend.text = element_text(size = 15),
    axis.text = element_text(size = 15),
    axis.title = element_text(size = 15)
)

fraction_sig_assoc_plot

```

# Factor-feature scatter plots

```{r factor_feature_plots, echo=FALSE}

print("Factor 3, virome")

plot_data_scatter(
    model,
    view = "virome",
    factor = 2,
    features = 15,
    add_lm = TRUE
)

```

```{r save_plots, echo=FALSE, eval=save_data}

# Fraction of significant associations

ggsave(
    filename = file.path(
        plot_path,
        "manuscript",
        "factor_feature_corr",
        "fraction_sig_associations.png"
    ),
    plot = fraction_sig_assoc_plot,
    width = 10,
    height = 10,
    dpi = 300
)

```

```{r save_plot_data, echo=FALSE, eval=save_data}

# Fraction of significant associations

write.csv(
    frac_sig_associations,
    file.path(
        plot_path,
        "manuscript",
        "plot_data",
        "fraction_sig_associations.csv"
    ),
    row.names = FALSE
)

# Factor-feature correlations

factor_feature_corr_table <- data.frame(
    corr_coefs,
    p_vals,
    p_vals_adjusted
)

write.csv(
    factor_feature_corr_table,
    file.path(
        plot_path,
        "manuscript",
        "plot_data",
        "factor_feature_corr.csv"
    ),
    row.names = TRUE
)

```

```{r,eval=TRUE}
sessionInfo()
```
